## IS605 Assignment 5
Paul Garaud

### PS 1
```{r}
# Ax = b
A <- matrix(c(1, 1, 1, 1, 0, 1, 3, 4), nrow=4)
b <- c(0, 8, 8, 20)
```

* Compute $A^TA$ and $A^Tb$
```{r}
AA <- t(A) %*% A
Ab <- t(A) %*% b
```

* Solve for $\hat{x}$ in R using the above two computed matrices
```{r}
(x.hat <- solve(AA) %*% Ab)
```

* What is the squared error of this solution?
```{r}
(e2 <- t(Ab - AA %*% x.hat) %*% (Ab - AA %*% x.hat))
```

* Instead of $b$, start with $p = \begin{bmatrix} 1 \\ 5 \\ 13 \\ 17 \end{bmatrix}$

Solution to Ax = p is $x = \begin{bmatrix} 1 \\ 4 \end{bmatrix}$ since:
```{r}
p <- c(1, 5, 13, 17)
Ap <- t(A) %*% p
x <- A %*% c(1, 4)
all.equal(x, as.matrix(p))  # test equivalence of every element

# Show that error vector e = 0
(Ap - t(A) %*% x)  # e = A'p - A'x
```

* Show that the error $e = b - p = \begin{bmatrix} -1 \\ 3 \\ -5 \\ 3 \end{bmatrix}$
```{r}
e.bp <- b - p
all.equal(e.bp, c(-1, 3, -5, 3))
```

* Show that $e$ is orthogonal to $p$ and each column of $A$
```{r}
(e.bp %*% p)
apply(A, 2, function(col) {e.bp %*% col})  # multiply e.bp and each col of A
```

### PS 2

```{r}
# load data
fname <- 'auto-mpg.data'
if (!(fname %in% dir())) {
  fname <- choose.files(default = fname, caption = 'Please select *.data file.')
}
auto <- read.table(fname)

# create A matrix and b vector
A <- as.matrix(auto[, 1:4])
b <- as.matrix(auto[, 5])

# Least Squares fit (no intercept term)
(coef <- solve(t(A) %*% A) %*% t(A) %*% b)
```
Thus the best fitting linear model is:

$\hat{mpg} = -0.03V_1 + 0.157V_2 - 0.006V_3 + 1.997V_4$

The error between predicted mpg, $\hat{mpg}$, and actual $mpg$:

```{r}
mpg.pred <- A %*% coef
t(b - mpg.pred) %*% (b - mpg.pred)  # squared sum of residuals
```