# IS605 Assignment 9
Paul Garaud

## PS 1

(1) Write a function to produce a sample of a random variable distributed as:

$$
f(x) = x, \hspace{.2cm} 0 \le x \le 1 \\
f(x) = 2 - x, \hspace{.2cm} 1 < x \le 2
$$

```{r step1}
dist1 <- function(y) {
  ifelse(y < 0.5, sqrt(2 * y), 2 - sqrt(2 - 2 * y))
}
```

(2) Write a function to produce a sample of a random variable distributed as:

$$
f(x) = 1 - x, \hspace{.2cm} 0 \le x \le 1 \\
f(x) = x - 1, \hspace{.2cm} 1 < x \le 2
$$

```{r step2}
dist2 <- function(y) {
  ifelse(y < 0.5, 1 - sqrt(1 - 2 * y), 1 + sqrt(2 * y - 1))
}
```

(3) Draw 1000 samples from both distributions & plot the histograms.

```{r step3}
hist(sapply(runif(1000), dist1), freq = FALSE, main='dist1', xlab='x')
hist(sapply(runif(1000), dist2), freq = FALSE, main='dist2', xlab='x')
```

(4) Write a program that takes sample size $n$ and a PDF function as parameters. The program will take $n$ samples & compute the mean of these samples, repeating this process 1000 times, before plotting these means.

```{r step4}
plot.means <- function(n, pdf) {
  hist.norm(
    sapply(
      1:1000, function(x) mean(sapply(runif(n), pdf))
    )
  )
}
```

(5) Verify that setting $n$ equal to a small number (10-20) still results in normally distributed means of the samples for each of the two PDF's to empirically verify the validity of the Central Limit Theorem.

```{r step5, message=FALSE}
require(ggplot2)
# helper function to plot histogram with estimated normal distribution
hist.norm <- function(data) {
  df <- data.frame(x = data)
  ggplot(data = df, aes(x = x)) + geom_histogram(aes(y = ..density..)) +
    stat_function(fun = dnorm, 
                  args = list(mean = mean(df$x), sd = sd(df$x)),
                  color = 'red', size = .75) +
    theme_bw()
}

plot.means(10, dist1)
plot.means(10, dist2)
```