# IS605 Assignment 10
Paul Garaud

## PS 1
Form the $A$ matrix and introduce decay (B matrix).

```{r part1}
A <- matrix(c(
  0,    0,    0.25, 0,    0,    0,
  0.5,  0,    0.25, 0,    0,    0,
  0.5,  1,    0,    0,    0,    0.5,
  0,    0,    0,    0,    0.5,  0.5,
  0,    0,    0.25, 0.5,  0,    0,
  0,    0,    0.25, 0.5,  0.5,  0
  ), nrow=6, byrow=TRUE
)
B <- 0.85 * A + (0.15 / nrow(A))  # cols should sum to 1, entries to 6
```

Start with a uniform rank vector $r$ and perform power iterations on B until convergence.

```{r part2}
r <- rep(1/6, 6)
C <- B
for (i in 1:20) {
  C <- C %*% C
}
r.final <- as.numeric(C %*% r)
r.length <- sqrt(sum(r.final**2))
```

Compute the eigen-decomposition of B and verify that the largest eigenvalue is 1 and that the corresponding eigenvector is the same as the final result above, has all positive entries, and sums to 1.

```{r part3}
B.eig <- eigen(B)
as.numeric(B.eig$values[1])  # largest eigenvalue is 1
eig.vect <- abs(B.eig$vectors[, 1])

# since unit vector, we need to scale by length of r.length
r.eig <- eig.vect * r.length
all.equal(r.eig, r.final)  # |eig vector| = r.final.unit
r.eig > 0  # all entries > 0
sum(r.eig)  # sums to 1
```

Use the `graph` package in R and its `page.rank` method to compute the Page Rank of the graph as given in A. Verify that results are analogous to above.

```{r part4, message=FALSE}
library(igraph)
g <- graph(c(1, 2, 1, 3, 2, 3, 3, 1, 3, 2, 3, 6, 6, 3, 3, 5,
             5, 6, 6, 4, 4, 6, 5, 4, 4, 5), directed = TRUE)
plot(g)  # verify that above graph matches notes
ranks <- page.rank(g)$vector
all.equal(r.final, ranks)
```